---
title: "Example Analysis"
format: html
---
# Economic and Health Burden of Non-Fatal Injuries 

The Global Burden of Diseases, Injuries, and Risk Factors Study (2021)  ?? idk say more here. 

This analysis uses data from the CDC Web-based Injury Statistics Query and Reporting System (WISQARS) [reference] on the number of cases and economic costs associated with various mechanisms of non-fatal injuries in 2023. The audience for this analysis are students, researchers, and those who are interested in understanding the magnitude of the economic and health burden of non-fatal injuries in the United States. 

## Data Dictionary



## Questions of interest

1. Which mechanisms of non-fatal injury were the most common in 2023? 
2. How does the mechanism of non-fatal injury differ by gender in 2023? 
3. 

**Loading packages**
```{r}
library(readr)
library(tidyverse)
library(dplyr)
```

**Loading data**
```{r}
costs <- read_csv("cost-data-export.csv")
cases <- read_csv("reports-data-export.csv")
```

